In-silico molecular enrichment and clearance of the human intracranial space

### 2024-12-19 paper/cut.tex
Move cut text from Results 2 and 1 to cut.tex and map out word count.
Clean-up in peristalsis in Methods

>>>>

\section{Notes}

Various notes associated with \cite{eide2024functional}:
\begin{itemize}
  \item
    Can observations of tracer indicate compartmentalization without
    the existence of physical barriers but in the presence of
    e.g.~substantial flow? We made observations in this direction in
    \cite{vinje2021brain}, relating to the then-active
    shape-of-perivascular-spaces discussion. We can address this by
    trying to quantify and evaluate the permeability of the PVS-SAS
    interface - comparing model predictions with the timings and
    patterns presented by \cite{eide2024functional}.
  \item
    Do the tracer primarily move from the PVS into the SAS or along
    (more slowly) along the SAS? This is left as an open question by
    \cite{eide2024functional} (page 3)
  \item
    How do altered ICP affect transport? Note that PVSAS transport is
    impaired (delayed) with increasing mean wave amplitude in the ICP
    signal and for iNPH patients (with enlarged PVSAS spaces).
  \item
    Very useful reference for PVS sizes (10--20 mm$^2$ in reference
    patients, 15-70 mm$^2$ in iNPH), and average propagation speeds.
\end{itemize}

%Flow of CSF through the SAS and the ventricular system is a crucial driver of transport in the human brain environment.
%While cardiac- and respiratory induced oscillatory flow increases mixing, the production of CSF in the choroid plexus creates a steady advective flow within the CSF-filled spaces of the cranium \cite{hornkjol2022csf}. 
%We construct a 3D model of the SAS and the ventricular system (\Cref{fig:csf}A,B and C) and compute CSF flow fields for three different scenarios: CSF production in the choroid plexus, CSF flow at systolic peak blood inflow and CSF flow in response to free breathing.

% production flow simulation
%% To simulate the production-driven flow, we impose an inflow of $400\,$ml/day across the surface of the lateral ventricle and allow a resistive outflow across the upper convexity of the arachnoid mater, leading to a total pressure drop of $26\,$mPa between the production and outflow sites (\Cref{fig:csf}D). A maximum flow velocity of $1.8\,$mm/s is attained in the aqueduct, while the mean velocity in the SAS is $2.6\,\mu$m/s (\Cref{fig:csf} E).

%% % cardiac pulsatility
%% Assessing the effect of cardiac-driven dispersion, we compute a steady CSF flow field at peak systolic blood inflow. We mimic the resulting reduction of CSF space volume by imposing an inflow of $6\,$ml/s \cite{baledent2014imaging, causemann2022human} and $0.31\,$ml/s \cite{vinje2019respiratory} across the upper arachnoid mater ($\Gamma_{\rm AM-U}$) and the lateral ventricle ($\Gamma_{\rm VL}$) surface, respectively. We allow free outflow into the spinal SAS ($\Gamma_{\rm SSAS}$). In addition, we upscale the obtained pressure field to account for the lack of pulsatile inertial effects in our model (see \Cref{sec:csf_dispersion}). In this scenario, we observe a pressure drop of $10\,$Pa between lateral ventricles and the spinal SAS and a maximum flow velocity of $19.8\,$cm/s in the aqueduct (\Cref{fig:csf}F and J).
%% Next, we estimate a spatially varying dispersion enhancement factor $R$ from cardiac-induced pulsatility by combining the computed pressure field with the theory of shear-augmented dispersion (see \Cref{sec:csf_dispersion}). Assuming a cardiac frequency of $1\,$Hz, we find that cardiac-induced pulsatile CSF flow increases the effective diffusion by more than two orders of magnitude in the aqueduct and around the cisterna magna, but has little effect in most of the SAS ($R < 1$) (\Cref{fig:csf} G and K).

%% % respiratory pulsatility
%% Employing the same methodology, we estimate the respiratory contribution to dispersive mixing by computing the CSF flow field driven by an inflow of $0.121\,$ml/s \cite{liu2024using} and $1\,$ml/s \cite{gutierrez2022effect} across the lateral ventricular wall and the arachnoid mater, respectively. This amounts to a respiratory peak flow volume of $1.121\,$ ml/s at the craniocervical junction. While the resulting flow velocities are only about one fourth of their cardiac-induced counterparts (see \Cref{fig:csf}H and L), respiratory dispersive mixing reaches a factor of up to 350 due the lower respiratory frequency of $0.25\,$Hz (\Cref{fig:csf}I and M).

%% % tracer trajectory of the standard model
%% \mer{MER: I moved this to Results 1 and condensed it a little bit}
%% \mar{Next, we use the computed advective velocity field and dispersion enhancement factors to predict the spreading of $0.5\,$mmol intrathecally injected Gadobutrol in CSF-filled spaces, the parenchymal tissue and PVSs around arteries and veines. We assume pressure-driven flow in arterial and venous PVSs, and cardiac peristaltic flow in the arterial network (\Cref{sec:pvs_flow_results}). One hour post-injection, tracer moves upwards in the SAS frontally of the brainstem,  and quickly reaches supratentorial regions (\Cref{fig:csf}N). Here, it spreads both posteriorly through quadrigeminal cistern and longitudinal fissure, and anteriorly through the outer SAS, reaching the top of the cerebral cortex after around 12 h. After 24h, the tracer covers most of the brains surface (with exception of some posterior regions) and has penetrated several centimeters into the parenchymal tissue.
%% This pattern of transport is reflected in the mean tracer concentrations in each compartment, where the CSF space reaches its peak of $1.4\,$mmol/l after 2 hours, followed by the the arterial PVS concentration peak ($2.9\,$mmol/l after 4 hours) (\Cref{fig:csf}P).  While the mean concentrations in these compartments fall quickly soon after their peaks, the parenchymal tissue slowly enriches with tracer over the first 24 hours, with a final value of $0.3\,$ mmol/l.
%% Overall, about 40\,\% of the total amount of tracer remains in the cranium 24\,h post-injection, with the largest share in the CSF (53\,\%), followed by the parenchyma (35\,\%), the venous PVS (8\,\%) and the arterial PVS (5\,\%). (\cref{fig:csf}O).}
